Building a continuous benchmarking ecosystem in bioinformatics

### 2025-07-23 2_second_review/manuscript_with_tracked_changes.tex
accept all changes, before replacing to bibitems

\author[1,2,*]{Izaskun Mallona}
\author[2,3]{Charlotte Soneson}
\author[1]{Ben Carrillo}
\author[1,2,4]{Almut Lütge}
\author[1]{Daniel Incicau}
\author[1,2]{Reto Gerber}
\author[1,2]{Anthony Sonrel}
\author[1,2,*]{Mark D. Robinson}

@@

Currently, most benchmarking systems aim to enhance the workflow formalization and execution with strategies to fulfill benchmarking-specific needs. To name a few of these systems, \texttt{ncbench} bundles workflow specification and software management with \texttt{snakemake} to benchmark result visualization with \texttt{Datavzrd} \cite{Hanssen2023-uf}. \texttt{OpenEBench} runs workflows in any workflow language, frequently \texttt{nextflow}, and makes use of \texttt{openVRE} as GUI and means of visualization \cite{Capella-Gutierrez2017-dh}. \texttt{OpenProblems} uses \texttt{Viash} \cite{cannoodt2024viash} and \texttt{nextflow} to handle software environments and workflows, respectively, and reports updated leaderboards with custom code \cite{luecken2025open}. To our knowledge, no benchmarking platform makes use of a fully expressive benchmark definition formalization language covering workflow generation and benchmarking-specific tasks. {\color{red} Orthogonally, some components of the benchmarking process have been formalized: \texttt{Viash} \cite{cannoodt2024viash} systematizes workflow components; \texttt{PROV-O} represents provenance data \cite{LeboUnknown-om}; and \texttt{Research Objects} (\texttt{RO} \cite{bechhofer2013linked}, e.g., \texttt{RO-Crate} \cite{soiland2022packaging}) specify a system to store data (files) and their metadata. \sout{Such a language would go a long way toward facilitating benchmark conceptualization, community gathering and result sharing} Reusing pre-existing formalizations and extending them to the benchmarking field would facilitate conceptualization, community gathering and result sharing}.

@@

{\color{red}Compute-intensive benchmarks are often run using cloud or high performance computing, further extending the challenge to ensure hardware homogeneity (for comparable performances) and adding complexity to the benchmarking system capabilities to automate node provisioning (or node selection) and to group executions such that costs are minimized. Some workflow managers, such as \texttt{Snakemake}, provide plugins to defer execution to cloud computing or queuing systems, effectively decoupling workflow specification and execution.}

@@

{\color{red} Authorship, e.g., acknowledging contributions, can be captured, reported and credited using third-party platforms such as Apicuron \cite{hatos2021apicuron} or by bundling standardized contribution metadata such as the Citation File Format (\texttt{CFF}). In this way, individual contributions to a benchmark can be acknowledged, browsed, and ideally propagated to academic deliverables (e.g., datasets, publications).}

@@

{\color{red} Transparency and clear governance guidelines provide an effective means to detect and react to potential misuse, e.g., by explicit rules to validate results. Other ethical considerations include facilitating safe interactions and fair resource allocation and efficiency. Clear code of conduct statements proactively flagging harassment (e.g., by sex, ethnicity or academic seniority, among others) and provide an explicit means to ease collaboration. In terms of efficiency, comparable allocation of resources (e.g., computing power, support) to all contributors, while also being aware of $CO_2$ emissions, could also be spelled out in the governance model.}

@@

Software, both from benchmarking systems or from the benchmarking workflow and hence benchmarked methods, can be run in different hardware, from commodity computers to dedicated hardware, often large servers or academic and commercial clouds. Performance profiling (e.g., how fast a benchmarked method runs) requires access to fair and comparable resources.

@@

Benchmarking is a cornerstone practice in bioinformatics, impacting how methods are conceived, developed, published and chosen. Benchmarking is driven by the effort of people with the expertise to address biological and technical problems. This expertise can be leveraged by current practices of community benchmarking, so scientific advances are eased by joint efforts to catalog open problems, generate (or simulate) appropriate input data, develop methods to address them, and evaluate their performances. Community benchmarking has been proven successful and  productive \cite{moult2005-ne,Capella-Gutierrez2017-dh,luecken2025open}}.

>>>>

\author[1,2,*]{Izaskun Mallona}
\author[2,3]{Charlotte Soneson}
\author[1,2]{Ben Carrillo}
\author[1,2,4]{Almut Lütge}
\author[1]{Daniel Incicau}
\author[1,2]{Reto Gerber}
\author[1,2]{Anthony Sonrel}
\author[1,2,*]{Mark D. Robinson}

@@

Currently, most benchmarking systems aim to enhance the workflow formalization and execution with strategies to fulfill benchmarking-specific needs. To name a few of these systems, \texttt{ncbench} bundles workflow specification and software management with \texttt{Snakemake} to benchmark result visualization with \texttt{Datavzrd} \cite{Hanssen2023-uf}. \texttt{OpenEBench} runs workflows in any workflow language, frequently \texttt{nextflow}, and makes use of \texttt{openVRE} as GUI and means of visualization \cite{Capella-Gutierrez2017-dh}. \texttt{OpenProblems} uses \texttt{Viash} \cite{cannoodt2024viash} and \texttt{nextflow} to handle software environments and workflows, respectively, and reports updated leaderboards with custom code \cite{luecken2025open}. To our knowledge, no benchmarking platform makes use of a fully expressive benchmark definition formalization language covering workflow generation and benchmarking-specific tasks. {\color{red} Orthogonally, some components of the benchmarking process have been formalized: \texttt{Viash} \cite{cannoodt2024viash} systematizes workflow components; \texttt{PROV-O} represents provenance data \cite{LeboUnknown-om}; and \texttt{Research Objects} (\texttt{RO} \cite{bechhofer2013linked}, e.g., \texttt{RO-Crate} \cite{soiland2022packaging}) specify a system to store data (files) and their metadata. \sout{Such a language would go a long way toward facilitating benchmark conceptualization, community gathering and result sharing} Reusing pre-existing formalizations and extending them to the benchmarking field would facilitate conceptualization, community gathering and result sharing}.

@@

{\color{red}Compute-intensive benchmarks are often run using cloud or high performance computing, further extending the challenge to ensure hardware homogeneity (for comparable performances) and adding complexity to the benchmarking system capabilities to automate node provisioning (or node selection) and to group executions such that costs are minimized. Some workflow managers, such as \texttt{Snakemake}, provide plugins to execute on the cloud or on-premises queuing systems, effectively decoupling workflow specification and execution.}

@@

{\color{red} Authorship, e.g., acknowledging contributions, can be captured, reported and credited using third-party platforms such as Apicuron \cite{hatos2021apicuron} or by bundling standardized contribution metadata such as the Citation File Format (\texttt{CFF}) \cite{druskat_2021_5171937}. In this way, individual contributions to a benchmark can be acknowledged, browsed, and ideally propagated to academic deliverables (e.g., datasets, publications).}

@@

{\color{red} Transparency and clear governance guidelines provide an effective means to detect and react to potential misuse, e.g., by explicit rules to validate results. Other ethical considerations include facilitating safe interactions and fair resource allocation and efficiency. Clear code of conduct statements proactively flagging harassment (e.g., by sex, ethnicity or academic seniority, among others) and provide explicit means to ease collaboration. In terms of efficiency, comparable allocation of resources (e.g., computing power, support) to all contributors, while also being aware of $CO_2$ emissions, could also be spelled out in the governance model.}

@@

Software, both from benchmarking systems and from the benchmarking workflow itself (the benchmarked methods), can be run on different hardware, from commodity computers to dedicated hardware, often large servers or academic and commercial clouds. Performance profiling (e.g., how fast a benchmarked method runs) requires access to fair and comparable resources.

@@

Benchmarking is a cornerstone practice in bioinformatics, impacting how methods are conceived, developed, published and chosen. Benchmarking is driven by the effort of people with the expertise to address biological and technical problems. This expertise can be leveraged by current practices of community benchmarking, so scientific advances are eased by joint efforts to catalog open problems, generate (or simulate) appropriate input data, develop methods to address them, and evaluate their performances. Community benchmarking has been proven successful and  productive \cite{moult2005-ne,Capella-Gutierrez2017-dh,luecken2025open}.
